---
title: "Blogdown"
author: "Yiluan Song"
date: "2023-09-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Read blogdown instructions [here](https://bookdown.org/yihui/blogdown/).

## Set up a blogdown project
Create an R project. It is strongly recommended to make the working directory a GitHub repo for version control and deployment.

Make this folder a blogdown site using R package _blogdown_.
```{r, eval=F}
install.packages("blogdown")
blogdown::new_site()
```

HUGO may be installed automatically, but manually install HUGO if necessary.
The example below is for windows. See full installation instructions [here](https://gohugo.io/installation/).
```{bash, eval = F}
winget install Hugo.Hugo.Extended
```

If the installed HUGO cannot be found, set PATH to include the path for HUGO
```{r, eval = F}
new_path <- "[path where HUGO is installed]"
current_path <- Sys.getenv("PATH")
new_path_list <- c(current_path, new_path)
Sys.setenv(PATH = paste(new_path_list, collapse = ";"))
Sys.getenv("PATH")
```

Your working directory should have new folders and files used for the blogdown site.

A sample site can be previewed in the Viewer.

Stop and serve your site to preview when you make further changes.
```{r, eval = F}
blogdown::stop_server()
blogdown::serve_site()
```

## Change blog content
Create a new R markdown for your new blog post.
Save it as content/post/[YYYY]-[MM]-[DD]-[title]/index.Rmd following the other examples.

Delete the other example blog post folders.

If there are files needed to knit this R markdown, save the files in the same folder as the markdown.

## Customize website elements
The customization below applies to the theme "lithium," but other themes might be customized in a similar fashion.

Change content in "About" page, found in content/about.md.

Change GitHub link to the link of this project in config.yaml.

Change Twitter link also in config.yaml. You can also delete this menu item or add other ones.

Change logo (also the "Home" button). Save your desired logo at themes/hugo-lithium/static/images/, where the original logo is saved. Change config.yaml params: logo: url: to the new file name.

Change favicon (tab preview image). Save your desired logo at themes/hugo-lithium/static/, where the original favicon is saved. Change config.yaml params: favicon: to the new file name.

Change website title in config.yaml title:.

Credit: Logo of this site was downloaded [here](https://www.pngitem.com/middle/obxRRi_otter-otter-icon-png-transparent-png/).

## Deploy site as a GitHub page
You can deploy either on GitHub or Netlify. Only GitHub deployment is discussed here.

Build site locally.
```{r, eval = F}
blogdown::build_site()
```
By default, the built webpages are stored in public/, but we want to change the directory name to docs/ for GitHub deployment. To do this, add the line of "publishDir: docs" in config.yaml. Build site again. Don't forget to delete public/.

Change baseurl from relative url to absolute url of the GitHub page, otherwise GitHub page does not get the paths right. To do this, change baseurl: in line 1 of config.yaml from "/" to "https://[username].github.io/[repo name]/".

To avoid some bugs in navigation using the "Home" button, I made changes to the  themes/hugo-lithium/layouts/partials/nav.html file. I changed the {{ "/" | relURL }} in line 2 to {{ .Site.BaseURL }}.

To fix website title display, I made changes to the themes/hugo-lithium/layouts/partials/head.html file in two places. I changed both {{ if eq .RelPermalink "/" }} in {{ if eq .Permalink .Site.BaseURL }}.

To make sure all content in docs/ are pushed to GitHub, comment out docs/ in .gitignore if applicable.
Push changes to GitHub. Remember to build site again before pushing.

Go to repo on the GitHub website. Settings > Pages > Build and deployment. Select main branch (or whichever you want), /docs, and save.

The website should deploy automatically upon pushing.
If deployed successfully, there should be a green tick next to the commit hash. Otherwise, there is a red cross.
Click on the red cross or green tick to inspect more details about the deployment.

The website url should be "https://[username].github.io/[repo name]/". You can also find the link in the last step of deployment details.
